---
title: "Introduction to the tidyverse in R"
author: "Jean V. Adams"
date: "Table of Contents"
output:
  html_document:
    toc: TRUE
    toc_depth: 3
    number_sections: false
    theme: cerulean
    highlight: tango
subtitle: USGS GLSC/LSC All Hands Meeting 2020
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, eval=FALSE, warning=FALSE, message=FALSE)
```

# Purpose

This session will introduce participants to general data wrangling using [tidyverse](https://www.tidyverse.org/) functions in R, including rearranging, subsetting, and summarizing data using a "piped" workflow.

This presentation and all the code is available at my GitHub site, in the **vizzy** repository, [**GitHub.com/JVAdams/vizzy**](https://github.com/JVAdams/vizzy).

This tutorial closely follows Garrett Grolemund and Hadley Wickham's 2017 book, [**R for Data Science**](https://r4ds.had.co.nz) (**R4DS**), freely available on-line.

# What you need

* [Download](https://cran.r-project.org/) and install R
* Install and load the tidyverse and nycflights13 packages

```{r, eval=FALSE}
install.packages("tidyverse")
install.packages("nycflights13")

library(tidyverse)
library(nycflights13)
```

```{r, echo=FALSE}
library(tidyverse)
library(nycflights13)
```

# Look at the data

We'll use the `flights` data with information on all 336,776 flights that departed from New York City in 2013.

```{r}
flights
```

Read about the fields.

```{r, eval=FALSE}
?flights
```

If you're using RStudio, browse the data.

```{r, eval=FALSE}
View(flights)
```

#	Data manipulation

## **filter()** - subset rows

The first argument is the name of the data frame. 

The remaining arguments are expressions that filter the data frame using comparison operators.

For example ...

Find all flights on January 1st:

```{r}
filter(flights, month == 1, day == 1)
```

Find all flights that departed in March, November, or December:

```{r}
filter(flights, month == 3 | month > 10)
```

For more info see **R4DS** [Filter rows with filter()](https://r4ds.had.co.nz/transform.html#filter-rows-with-filter).

## **arrange()** - reorder rows

The first argument is the name of the data frame. 

The remaining arguments are columns to sort by.

For example ...

Sort all the flights by date:

```{r}
arrange(flights, year, month, day)
```

Use `desc()` to order by a column in descending order.

For example ...

Sort by departure delays (`dep_delay`), with the longest delays first:

```{r}
arrange(flights, desc(dep_delay))
```

For more info see **R4DS** [Arrange rows with arrange()](https://r4ds.had.co.nz/transform.html#arrange-rows-with-arrange).

## **select()** - subset variables

The first argument is the name of the data frame. 

The remaining arguments are columns to keep.

For example ...

Subset the data frame with just the date columns:

```{r}
select(flights, year, month, day)
```

You can use a colon (:) to select a range of variables.

For example ... 

Select all columns between year and departure time (`dep_time`):

```{r}
select(flights, year:dep_time)
```

For more info see **R4DS** [Select columns with select()](https://r4ds.had.co.nz/transform.html#select).

## **mutate()** - create new variables

The first argument is the name of the data frame. 

The remaining arguments are new columns to create.

For example ...

Create a new variable called `speed` (in miles per hour) calculated from `distance` (in miles) and `air_time` (in minutes):

```{r}
mutate(flights, 
  speed = 60 * distance / air_time
)
```

We've got too many variables to see the new variable we just created.  Save the data frame you just created, then subset it to just a few variables:

```{r}
dat1 <- mutate(flights, 
  speed = 60 * distance / air_time
)
select(dat1, origin, dest, distance, air_time, speed)
```

You can create multiple variables at once, and you can refer to columns that you just created:

```{r}
dat1 <- mutate(flights, 
  air_hours = air_time / 60,
  speed = distance / air_hours
  )
select(dat1, origin, dest, distance, air_hours, speed)
```

For more info see **R4DS** [Add new variables with mutate()](https://r4ds.had.co.nz/transform.html#add-new-variables-with-mutate).

## **%>%** - pipes

Why pipes?  Pipes help you clearly express a sequence of multiple operations.

Let's return to our previous example, and rewrite the code with pipes.

```{r}
flights %>% 
  mutate(
    air_hours = air_time / 60,
    speed = distance / air_hours
  ) %>%
  select(origin, dest, distance, air_hours, speed)
```

The pipe operator, `%>%`, basically says, take the data on the left and apply the function on the right.  

It assumes that the function's first argument is the data frame.  *Conveniently, **all** of the functions that we've been talking about take a data frame as their first argument.*  

It allows us to string together a bunch of operations without having to create temporary data frames (like `dat1`) in between.

Often, when applying a bunch of functions, you'll want to save the result to a different data frame.  That's easy.  Just start off with the name of the new data frame.

```{r}
speed.dat <- flights %>% 
  mutate(
    air_hours = air_time / 60,
    speed = distance / air_hours
    ) %>%
  select(origin, dest, distance, air_hours, speed)
```

For more info see **R4DS** [Pipes](https://r4ds.had.co.nz/pipes.html).

### Your turn

Use pipes to find the fastest flights on January 1st:

 * find all flights on January 1st,
 * calculate `speed`,
 * sort by `speed` (with fastest first), and
 * select destination (`dest`), `distance`, `air_time`, and `speed`.

```{r, echo=FALSE}
flights %>%
  filter(month == 1, day == 1) %>%
  mutate(speed = 60 * distance / air_time) %>%
  arrange(desc(speed)) %>%
  select(dest, distance, air_time, speed)
```

## **summarise()** - collapse data to a single row

The first argument is the name of the data frame (sound familiar?).

The remaining arguments are summary functions.

For example ...

Calculate the mean departure delay:

```{r}
summarise(flights, 
  delay = mean(dep_delay, na.rm = TRUE)
)
```

You can extend these summary calculations to multiple groups using the `group_by()` function. 

For example ...

Apply the exact same code above to the `flights` data after grouping by month:

```{r}
flights %>%
  group_by(month) %>%
  summarise(
    delay = mean(dep_delay, na.rm = TRUE)
  )
```

Now, add the number of flights to the summary:

```{r}
flights %>%
  group_by(month) %>%
  summarise(
    delay = mean(dep_delay, na.rm = TRUE),
    nflights = n()
  )
```

For more info see **R4DS** [Grouped summaries with summarise()](https://r4ds.had.co.nz/transform.html#grouped-summaries-with-summarise)



## **pivot_wider()**

Sometimes data are organized in long format for data analysis or plotting.  Often, you want to pivot the data into a wide format for easy viewing or tabulation.

The first argument is ... ?

The remaining arguments are

* `names_from`: the columns to take variable names from and
* `values_from`: the columns to take values from.

For example ...

First, create a new data frame that has the monthly mean departure delay for three airlines (`carrier`):

* AA - American,
* DL - Delta, and
* WN - Southwest.

```{r}
example <- flights %>%
  filter(carrier %in% c("AA", "DL", "WN")) %>%
  group_by(month, carrier) %>%
  summarise(
    delay = mean(dep_delay, na.rm = TRUE)
  )
```

Now, make a separate column of monthly delays for each airline.

```{r}
pivot_wider(example, 
  names_from = carrier, 
  values_from = delay
)
```

For more info see **R4DS** [Wider](https://r4ds.had.co.nz/tidy-data.html#wider).

## **pivot_longer()**

Sometimes data are organized in a wide format for ease of data entry or viewing.  Often, you want to pivot the data into a long format for data analysis or plotting.

The first argument is ... ?

The remaining arguments are

* `cols` - the columns that represent values,
* `names_to` - the name you want to assign to those values, and
* `values_to` - the name you want to assign to the current row values.

For example ...

Save the wide data we just created as a data frame.  Then convert it back to a long data format.

```{r}
wide.dat <- pivot_wider(example, 
  names_from = carrier, 
  values_from = delay
)

pivot_longer(wide.dat, 
  cols=AA:WN, 
  names_to = "airline", 
  values_to = "dep.delay.min"
)
```

For more info see **R4DS** [Longer](https://r4ds.had.co.nz/tidy-data.html#longer).


# Wrap up

## Summary

I've introduced you to several core functions for data manipulation in the tidyverse,

* `filter()` - subset rows
* `arrange()` - reorder rows
* `select()` - subset variables
* `mutate()` - create new variables
* `%>% ` - pipes
* `summarise()` - collapse data to a single row
* `pivot_wider()`
* `pivot_wider()`

## More info

Check out the [Cookbook for R](http://www.cookbook-r.com/) for more examples.

Stuck?  Try [Google](https://www.google.com/) search or search [Stackoverflow with R tags](https://stackoverflow.com/questions/tagged/r).

## Parting challenge

Create a new data frame from the `flights` data, containing information on the number of flights on the 27th `day` of each `month`, to six destinations (`dest`):

* `CUN` - Cancun,
* `ORD` - Chicago,
* `DTW` - Detroit,
* `IAD` - Dulles,
* `LAS` - Las Vegas, and
* `MIA` - Miami.

Sort the data frame by the number of flights, with the most flights first.

Then take this new data frame and make it wider, for easier reading as a table, and sort the data frame by month.

```{r, echo=FALSE}
challenge <- flights %>%
  filter(
    day == 27,
    dest %in% c("CUN", "ORD", "DTW", "IAD", "LAS", "MIA")
  ) %>%
  group_by(month, dest) %>%
  summarise(
    n.flights = n()
  ) %>%
  arrange(desc(n.flights))

challenge

pivot_wider(challenge, 
    names_from = dest, 
    values_from = n.flights
  ) %>%
  arrange(month)
```
