---
title: "Data table - tabular data display"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
load(url(huron.url))
```

The `datatable()` function from the **DT** package displays a data frame as an interactive table.  

Let's take a look at the bottom trawl operations data, using the defaults.

```{r, eval=FALSE}
library(DT)
datatable(optrop)
```

From this display, we can 

* change the number of entries displayed,
* sort by a column,
* search for a value, and
* page through the data.

```{r, echo=FALSE}
library(DT)
datatable(optrop)
```

# Reduce the number of rows displayed

```{r}
datatable(optrop, options=list(pageLength=5))
```

# Remove the row names

```{r}
datatable(optrop, rownames=FALSE)
```

# Show a subset of columns

```{r}
datatable(optrop[ , c("opid", "opdate", "depth", "fishtemp")])
```

# Add a caption

```{r}
datatable(optrop, caption="Lake Huron bottom trawl operations")
```

# Add the ability to filter

* Range sliders - numeric columns
* Select lists - factor (categorical) columns
* text input boxes - character columns

```{r, eval=FALSE}
datatable(optrop, filter="top")
```

For example, we can use the filter to look at the subset of data from years 2008-2009, port 308.  Or you could search by port name instead of port code.

```{r, echo=FALSE}
datatable(optrop, filter="top")
```

# Return to [main page](https://rawgit.com/JVAdams/vizzy/master/Visualizations.html)
