---
title: "Heatmap - interactive heatmap"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

internet <- TRUE
if(internet) {
  huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
  load(url(huron.url))
} else {
  huron <- "C:/JVA/GitHub/vizzy/Huron0610.RData"
  load(huron)
}
```

The `d3heatmap()` function from the **d3heatmap** package displays a heatmap with support for row and column highlighting and zooming.

First, we'll create a matrix of the average number of fish captured per operation by port-depth (in rows) and species (in columns).  If a given species was not captured at a given port and depth, We'll set its total to zero.

```{r}
totcat <- with(optrcat, 
  tapply(n, list(paste(port, format(depth)), species), mean))
totcat[is.na(totcat)] <- 0
```

Let's take a look at the matrix we created.

```{r}
library(DT)
datatable(totcat)
```

Now, let's make a heatmap from this matrix, using the defaults.

```{r, eval=FALSE}
library(d3heatmap)
d3heatmap(totcat)
```

The results are so skewed, only a few of the species are distinguishable from the rest.  

```{r, echo=FALSE}
library(d3heatmap)
d3heatmap(totcat)
```

Let's square root transform them.

```{r, eval=FALSE}
sqrttotcat <- sqrt(totcat)
d3heatmap(sqrttotcat)
```

You can see that the species columns and the port-depth rows have been re-ordered according to some default clustering algorithm.  The dendrograms are visible on the left side and top of the heatmap.  You can hover over a cell to find out what its row, column, and value are.  You can also zoom in on a selected set of cells.

```{r, echo=FALSE}
sqrttotcat <- sqrt(totcat)
d3heatmap(sqrttotcat)
```

# Increase font size

```{r}
d3heatmap(sqrttotcat, cexRow=2, cexCol=2)
```

# Change colors

```{r}
d3heatmap(sqrttotcat, colors="Blues")
```

# Keep the original column order

```{r}
d3heatmap(sqrttotcat, Colv=FALSE)
```

# Keep the original row order

```{r}
d3heatmap(sqrttotcat, Rowv=FALSE)
```

# Return to [main page](https://rawgit.com/JVAdams/vizzy/master/Visualizations.html)
