---
title: "Leaflet - geo-spatial mapping"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

internet <- TRUE
if(internet) {
  huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
  load(url(huron.url))
} else {
  huron <- "C:/JVA/GitHub/vizzy/Huron0610.RData"
  load(huron)
}
```

The `leaflet()` function from the **leaflet** package creates a dynamic map that supports panning and zooming.

Let's map the Lake Huron bottom trawl operations data, using the defaults.

The code that I'm using here makes use of a **piping** operator, **%>%**.  This operator essentially chains together a bunch of operations on a single object.  It's handy to use in this case, because the **leaflet** package builds a single map from a collection of separate functions.

After loading the package, 

* the `leaflet()` function identifies the data source,
* the `addTiles()` function creates the background map image, and
* the `addCircleMarkers()` function adds markers to the map.

```{r, eval=FALSE}
library(leaflet)
leaflet(optrop) %>% 
  addTiles() %>%
  addMarkers(lng=~long, lat=~lat)
```

We can pan and zoom our way around this map.

```{r, echo=FALSE}
library(leaflet)
leaflet(optrop) %>% 
  addTiles() %>%
  addMarkers(lng=~long, lat=~lat)
```

# Color the symbols by depth

In order to use different colors, we need to define a color palette.  In this case, we defined a palette of different shades of blue, corresponding to the bottom depth at each location. 

```{r}
numpal <- colorNumeric("Blues", NULL)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth))
```

# Add a legend

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth)) %>%
  addLegend(pal=numpal, values=~depth)
```

# Refine the legend

Change the position and transparency of the legend, and add a title.

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth)) %>%
  addLegend(pal=numpal, values=~depth,
    position="bottomleft", opacity=1, title="Bottom depth  (m)")
```

# Highlight one year

```{r}
facpal <- colorFactor(c("blue", "orange"), NULL)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~facpal(year==2010))
```

# Provide a toggle for that one year

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long[year!=2010], lat=~lat[year!=2010], 
    group="2006-2009") %>% 
  addCircleMarkers(lng=~long[year==2010], lat=~lat[year==2010], 
    group="2010") %>%   
  addLayersControl(overlayGroups=c("2006-2009", "2010"),
    options=layersControlOptions(collapsed=FALSE))
```

# Show the opid when you click on a point

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, popup=~as.character(opid))
```

# Use a different background map

```{r}
leaflet(optrop) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

```{r}
leaflet(optrop) %>% 
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

# Return to [main page](https://rawgit.com/JVAdams/vizzy/master/Visualizations.html)
