---
title: "Streamgraph - interactive stacked area plot"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
load(url(huron.url))
```

The `streamgraph()` function from the **streamgraph** package displays an interactive stacked area plot.

First, we'll create a data frame of the average number of fish captured per operation by year and species name.

```{r}
meannum <- aggregate(n ~ spname + year, data=optrcat, FUN=mean)
```

Let's take a look at the data frame we created.

```{r}
library(DT)
datatable(meannum)
```

Then we'll make a streamgraph from this data frame, using the defaults.  The arguments specify the data frame `data`, the category for the areas `key`, and the quantities for the x- `date` and y-axes `value`.

```{r}
library(streamgraph)
streamgraph(data=meannum, key="spname", date="year", value="n")
```

# Fix the x-axis

Change the tick interval to 1 year.  

The code that I'm using here makes use of a **piping** operator, **%>%**.  This operator essentially chains together a bunch of operations on a single object.  It's handy to use in this case, because the **streamgraph** package builds a single figure from a collection of separate functions.

```{r}
streamgraph(data=meannum, key="spname", date="year", value="n") %>%
  sg_axis_x(1, "year", "%Y")
```

# Make the image smaller

```{r}
streamgraph(data=meannum, key="spname", date="year", value="n",
  width=400, height=200) %>%
  sg_axis_x(1, "year", "%Y")
```

# Show as proportions (from 0 to 100%)

```{r}
streamgraph(data=meannum, key="spname", date="year", value="n",
  offset="expand") %>%
  sg_axis_x(1, "year", "%Y")
```

# Use a zero baseline

```{r}
streamgraph(data=meannum, key="spname", date="year", value="n",
  offset="zero") %>%
  sg_axis_x(1, "year", "%Y")
```

# Draw unsmoothed lines

```{r}
streamgraph(data=meannum, key="spname", date="year", value="n",
  interpolate="linear") %>%
  sg_axis_x(1, "year", "%Y")
```

# Draw rectangles

```{r}
streamgraph(data=meannum, key="spname", date="year", value="n",
  interpolate="step") %>%
  sg_axis_x(1, "year", "%Y")
```

# Return to [main page](https://rawgit.com/JVAdams/vizzy/master/Visualizations.html)
