---
title: "ScatterD3 - interactive scatter plot"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

internet <- TRUE
if(internet) {
  huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
  load(url(huron.url))
} else {
  huron <- "C:/JVA/GitHub/vizzy/Huron0610.RData"
  load(huron)
}
```

The `scatterD3()` function from the **scatterD3** package displays an interactive scatter plot.

First, let's create a simple scatter plot of fishing depth versus fishing temperature.  I use the `with()` function to refer to the data frame.

We can **zoom** (using the mouse scroll wheel), **pan** (by clicking and dragging), and we can **display** the plotted values (by hovering). 

```{r}
library(scatterD3)
with(optrop, scatterD3(depth, fishtemp))
```

# CAUTION: missing values

`scatterD3()` does weird things with missing values.  Let's create a data frame with no missing values for `fishtemp`.

```{r}
optrop.sub <- optrop[!is.na(optrop$fishtemp), ]

with(optrop.sub, scatterD3(depth, fishtemp))
```

# Add text labels

These can be repositioned.

```{r}
with(optrop.sub, scatterD3(depth, fishtemp, lab=pname))
```

# Use different colors

```{r}
with(optrop.sub, scatterD3(depth, fishtemp, col_var=year))
```

# Use different symbols

```{r}
with(optrop.sub, scatterD3(depth, fishtemp, symbol_var=cruise))
```

# Size the symbols

```{r}
with(optrop.sub, scatterD3(depth, fishtemp, size_var=surftemp))
```

# Specify text to display on hover

```{r}
with(optrop.sub, scatterD3(depth, fishtemp, tooltip_text=opid))
```

# Return to [main page](https://rawgit.com/JVAdams/vizzy/master/Visualizations.html)
