---
title: "Recent Advances in Interactive Visualizations in R"
author: "Jean V. Adams"
date: "26 Jan 2016"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
subtitle: USGS-GLSC All Hands Meeting
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```



# HTML widgets for R

HTML widgets produce interactive web visualizations

A few lines of R code is all it takes. 

I will walk through some examples of:

* Data table - interactive tabular data display   
* Streamgraph - interactive stacked area plot
* Heatmap - interactive heatmap (cluster grouping)  
* Leaflet - interactive geo-spatial mapping  

This presentation and all the code is available at my GitHub site, in the **vizzy** repository,

 * [**GitHub.com / JVAdams / vizzy**](https://github.com/JVAdams/vizzy)



# What you need

* [Download](https://cran.r-project.org/bin/windows/base/R-3.2.3-win.exe) and install R
* Install packages

```{r, eval=FALSE}
install.packages(c(
  "DT",
  "d3heatmap",
  "leaflet",
  "devtools"
)
devtools::install_github("hrbrmstr/streamgraph")
```

If you have difficulty downloading the **streamgraph** package from GitHub using the `install_github()` function, I suggest you download the zip file from the latest release at 

* [GitHub.com / hrbrmstr / streamgraph / releases](https://github.com/hrbrmstr/streamgraph/releases)

and install it from the R menu, 

* *Packages, Install package(s) from local zip files...*



# Read in the data

I will use a subset of five years of data from the Lake Huron bottom trawl survey to demonstrate the capabilities of HTML widgets for R.

```{r, echo=FALSE}
setwd("C:/JVA/GitHub/vizzy")
datadir <- "C:/JVA/Lakes Vessels/Huron/2012/"


sn <- read.csv(paste0(datadir, "Species.csv"), as.is=TRUE, 
  col.names=c("species", "spname", "sciname"))
# get rid of trailing white space
sn$spname <- gsub("^[ \t]+|[ \t]+$", "", sn$spname)
sn <- sn[, c("species", "spname")]


pn <- read.csv(paste0(datadir, "HuronPortNames.csv"), as.is=TRUE, 
  col.names=c("port", "pname"))


op <- read.csv(paste0(datadir, "HURON_OP.csv"), as.is=TRUE, 
  col.names=c("opid", "year", "vessel", "serial", "lake", "port", "cruise", 
    "opdate", "time", "lat", "long", "surftemp", "weather", "windspeed", 
    "seacond", "winddirect", "distance"))
numvars <- c("lat", "long", "surftemp", "weather", "windspeed", "seacond", 
  "winddirect", "distance")
op[, numvars] <- apply(op[, numvars], 2, as.numeric)
fixcoord <- function(ddmm., rnd=4) {
  dd <- floor(ddmm./100)
  mm. <- ddmm. - 100*dd
  dd. <- dd + mm./60
  dd. <- round(dd., rnd)
  return(dd.)
}
op$long <- -fixcoord(op$long)
op$lat <- fixcoord(op$lat)
op$opdate <- as.Date(op$opdate, "%d-%b-%y")
op$port <- as.factor(op$port)
op <- merge(op, pn)
sel <- !apply(is.na(op[, c("lat", "long")]), 1, any) & op$year %in% 2006:2010
op <- op[sel, c("opid", "year", "vessel", "cruise", "serial", "port", "pname", 
  "opdate", "time", "lat", "long", "surftemp")]


trop <- read.csv(paste0(datadir, "HURON_TROP.csv"), as.is=TRUE, 
  col.names=c("opid", "towtime", "fishtemp", "depth", "trdesign"))
trop$fishtemp <- as.numeric(trop$fishtemp)

optrop <- merge(op, trop[, c("opid", "towtime", "fishtemp", "depth")])
optrop$depth[with(optrop, depth > 60 & depth < 70)] <- 64


cat <- read.csv(paste0(datadir, "HURON_TRCATCH.csv"), as.is=TRUE, 
  col.names=c("opid", "stage", "species", "n", "weight", "lfn"))
numvars <- c("stage", "weight", "lfn")
cat[, numvars] <- apply(cat[, numvars], 2, as.numeric)
optrcat <- merge(optrop, cat)
optrcat <- optrcat[optrcat$species %in% 
    c("106", "108", "109", "130", "131", "133", "203", "204", "212", "307", 
      "5", "504", "508", "801", "803", "904", "910"), ]
optrcat <- merge(optrcat, sn)
optrcat$species <- as.factor(optrcat$species)
optrcat <- optrcat[, c("opid", "year", "port", "pname", "opdate", "lat", 
  "long", "towtime", "fishtemp", "depth", "species", "spname", "stage", "n", 
  "weight", "lfn")]
optrcat <- optrcat[with(optrcat, order(year, opdate, opid)), ]

save(optrop, optrcat, file="Huron0610.RData")
rm(list = ls(name = ".GlobalEnv"), pos = ".GlobalEnv")
```

If you wish to try out the examples I am demonstrating today, download the data from my GitHub site using the commands below.  

```{r}
huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
load(url(huron.url), verbose=TRUE)
```

The file contains two objects: the bottom trawl operations data, **optrop**, and catch data, **optrcat**.



# Data table - tabular data display  

# Streamgraph - interactive stacked area plot

# Heatmap - interactive heatmap

# Leaflet - geo-spatial mapping  



# Summary

We've looked at examples of the HTML widgets that I thought would be of the most use to you.

* Data table - interactive tabular data display   
* Streamgraph - interactive stacked area plot
* Heatmap - interactive heatmap (cluster grouping)  
* Leaflet - interactive geo-spatial mapping  

There are other HTML widgets available, too.  For more information, visit the HTML widgets [showcase](http://www.htmlwidgets.org/showcase_leaflet.html).
