---
title: "Recent Advances in Interactive Visualizations in R"
author: "Jean V. Adams"
date: "13 Jan 2016"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    number_sections: false
    theme: cerulean
    highlight: tango
subtitle: USGS-GLSC All Hands Meeting
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

# define two variables to keep track of Table and Figure numbers
tabcount <- 1
figcount <- 1
```



# HTML widgets for R

## HTML widgets produce interactive web visualizations

A few lines of R code is all it takes. 

I will walk through some examples of:

* Data table - interactive tabular data display   
* Heatmap - interactive heatmap (bivariate grouping)  
* Leaflet - interactive geo-spatial mapping  

This presentation is available at my GitHub site, in the **vizzy** repository,

 * [**github.com / JVAdams / vizzy**](https://github.com/JVAdams/vizzy)

## What you need:

* [Download](https://cran.r-project.org/bin/windows/base/R-3.2.3-win.exe) and install R
* Install packages

```{r, eval=FALSE}
install.packages(c(
  "DT",
  "d3heatmap",
  "leaflet"
)
```



# Read in the data

I will use a subset of five years of data from the Lake Huron bottom trawl survey to demonstrate the capabilities of HTML widgets for R.

```{r, echo=FALSE}
setwd("C:/JVA/GitHub/vizzy")
datadir <- "C:/JVA/Lakes Vessels/Huron/2012/"

op <- read.csv(paste0(datadir, "HURON_OP.csv"), as.is=TRUE, 
  col.names=c("opid", "year", "vessel", "serial", "lake", "port", "cruise", 
    "opdate", "time", "lat", "long", "surftemp", "weather", "windspeed", 
    "seacond", "winddirect", "distance"))
numvars <- c("lat", "long", "surftemp", "weather", "windspeed", "seacond", 
  "winddirect", "distance")
op[, numvars] <- apply(op[, numvars], 2, as.numeric)
fixcoord <- function(ddmm., rnd=4) {
  dd <- floor(ddmm./100)
  mm. <- ddmm. - 100*dd
  dd. <- dd + mm./60
  dd. <- round(dd., rnd)
  return(dd.)
}
op$long <- -fixcoord(op$long)
op$lat <- fixcoord(op$lat)
op$opdate <- as.Date(op$opdate, "%d-%b-%y")
sel <- !apply(is.na(op[, c("lat", "long")]), 1, any) & op$year %in% 2006:2010
op <- op[sel, c("opid", "year", "vessel", "cruise", "serial", "port", "opdate", "time", "lat", "long", "surftemp")]

trop <- read.csv(paste0(datadir, "HURON_TROP.csv"), as.is=TRUE, 
  col.names=c("opid", "towtime", "fishtemp", "depth", "trdesign"))
trop$fishtemp <- as.numeric(trop$fishtemp)

optrop <- merge(op, trop[, c("opid", "towtime", "fishtemp", "depth")])

cat <- read.csv(paste0(datadir, "HURON_TRCATCH.csv"), as.is=TRUE, 
  col.names=c("opid", "stage", "species", "n", "weight", "lfn"))
numvars <- c("stage", "weight", "lfn")
cat[, numvars] <- apply(cat[, numvars], 2, as.numeric)
optrcat <- merge(optrop, cat)

save(optrop, optrcat, file="Huron0610.RData")
rm(list = ls(name = ".GlobalEnv"), pos = ".GlobalEnv")
```

If you wish to try out the examples I am demonstrating today, download the data from my GitHub site using the commands below.  You will see that the file contains two objects: the bottom trawl operations data, **optrop**, and catch data, **optrcat**.

```{r}
huron.url <- "https://github.com/JVAdams/vizzy/raw/master/Huron0610.RData"
load(url(huron.url), verbose=TRUE)
```



# Data table - tabular data display  

The `datatable()` function from the **DT** package displays a data frame as an interactive HTML table.  

Let's take a look at the bottom trawl operations data, using the defaults.

From this display, we can 

* change the number of entries displayed,
* sort by a column,
* search for a value, and
* page through the data.

```{r}
library(DT)
datatable(optrop)
```

## Reduce the number of rows displayed

```{r}
datatable(optrop, options=list(pageLength=5))
```

## Remove the row names

```{r}
datatable(optrop, rownames=FALSE)
```

## Show a subset of columns

```{r}
datatable(optrop[ , c("opid", "opdate", "depth", "fishtemp")])
```

## Add a caption

```{r}
datatable(optrop, caption="Lake Huron bottom trawl operations")
```

## Add the ability to filter

For example, we can use the filter to look at the subset of data from port 308 in year 2008.

```{r}
datatable(optrop, filter="top")
```



# Heatmap - interactive heatmap

The `d3heatmap()` function from the **d3heatmap** package displays a heatmap with support for row and column highlighting and zooming.

First, we'll create a matrix of the average number of fish captured per operation by port-depth (in rows) and species (in columns).  If a given species was not captured at a given port and depth, We'll set its total to zero.

```{r}
totcat <- with(optrcat, tapply(n, list(paste(port, format(depth)), species), mean))
totcat[is.na(totcat)] <- 0
datatable(round(totcat))
```

Then we'll make a heatmap from this matrix, using the defaults.

```{r}
library(d3heatmap)
d3heatmap(totcat)
```

The results are so skewed, only a few of the species are distinguishable from the rest.  Let's square root transform them.

Let's take a closer look.  You can see that the species columns and the port-depth rows have been re-ordered according to some default clustering algorithm.  The dendrograms are visible on the left side and top of the heatmap.  You can hover over a cell to find out what its row, column, and value are.  You can also zoom in on a selected set of cells.

```{r}
sqrttotcat <- sqrt(totcat)
d3heatmap(sqrttotcat)
```

## Make it smaller

```{r}
d3heatmap(sqrttotcat, width=400, height=300)
```

## Increase font size

```{r}
d3heatmap(sqrttotcat, cexRow=2, cexCol=2)
```

## Change colors

```{r}
d3heatmap(sqrttotcat, colors="Blues")
```

## Keep the original column order

```{r}
d3heatmap(sqrttotcat, Colv=FALSE)
```

## Keep the original row order

```{r}
d3heatmap(sqrttotcat, Rowv=FALSE)
```



# Leaflet - geo-spatial mapping  

Leaflet creates a dynamic map that supports panning and zooming.

Let's map the Lake Huron bottom trawl operations data, using the defaults.

The code that I'm using here makes use of a **piping** operator, **%>%**.  This operator essentially chains together a bunch of operations on a single object.  It's handy here, because the **leaflet** package builds a single map from a collection of separate commands.

After loading the package, the first line uses the **leaflet()** function to idenitify the data source.  The second line uses the **addTiles()** function to create the background map image.  The third line uses the **addCircleMarkers()** function to add markers to the map at the location of bottom trawl operations.

We can pan and zoom our way around this map.

```{r}
library(leaflet)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

## Make the map smaller

```{r}
leaflet(optrop, width=500, height=300) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

## Color the symbols by depth

In order to use different colors, we need to define a color palette.  In this case, we defined a palette of different shades of blue, corresponding to the bottom depth at each location. 

```{r}
numpal <- colorNumeric("Blues", NULL)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth))
```

## Add a legend

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth)) %>%
  addLegend("bottomleft", pal=numpal, values=~depth, opacity=1,
    title="Bottom depth  (m)")
```

## Highlight one year

```{r}
facpal <- colorFactor(c("blue", "orange"), NULL)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~facpal(year==2010))
```

## Provide a toggle for that one year

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long[year!=2010], lat=~lat[year!=2010], 
    group="2006-2009") %>% 
  addCircleMarkers(lng=~long[year==2010], lat=~lat[year==2010], 
    group="2010") %>%   
  addLayersControl(overlayGroups=c("2006-2009", "2010"),
    options=layersControlOptions(collapsed=FALSE))
```

## Show the opid when you click on a point

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, popup=~as.character(opid))
```

## Use a different background map

```{r}
leaflet(optrop) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

# Summary

We've looked at examples of the HTML widgets that I thought would be of the most use to you.

* Data table - interactive tabular data display   
* Heatmap - interactive heatmap (colorized table)  
* Leaflet - interactive geo-spatial mapping  

There are other HTML widgets available, too.  For more information, visit the HTML widgets [showcase](http://www.htmlwidgets.org/showcase_leaflet.html).

