---
title: "Recent Advances in Interactive Visualizations in R"
author: "Jean V. Adams"
date: "11 Jan 2016"
output:
  html_document: default
subtitle: USGS-GLSC All Hands Meeting
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

# define two variables to keep track of Table and Figure numbers
tabcount <- 1
figcount <- 1
```

# **htmlwidgets** for R

> HTML widgets work just like R plots except they produce interactive web visualizations. 

> A line or two of R code is all it takes to produce a D3 graphic or Leaflet map. 

> HTML widgets can be used at the R console as well as embedded in R Markdown reports and Shiny web applications.

**DataTables** - Tabular data display   
**MetricsGraphics** - Scatterplots and line charts with D3  
**dygraphs** - Time series charting  
**Leaflet** - Geo-spatial mapping  
**networkD3** - Graph data visualization with D3  
**d3heatmap** - Interactive heatmaps with D3  

I will use historical data from the Lake Huron bottom trawl survey to demonstrate the capabilities of **htmlwidgets** for R.

For more information, visit the **htmlwidgets** showcase at http://www.htmlwidgets.org/showcase_networkD3.html.

# Read in the data

```{r, echo=FALSE}
setwd("C:/JVA/GitHub/vizzy")
datadir <- "C:/JVA/Lakes Vessels/Huron/2012/"
```

```{r}
op <- read.csv(paste0(datadir, "HURON_OP.csv"), as.is=TRUE, 
  col.names=c("opid", "year", "vessel", "serial", "lake", "port", "cruise", 
    "opdate", "time", "lat", "long", "surftemp", "weather", "windspeed", 
    "seacond", "winddirect", "distance"))
numvars <- c("lat", "long", "surftemp", "weather", "windspeed", "seacond", 
  "winddirect", "distance")
op[, numvars] <- apply(op[, numvars], 2, as.numeric)
fixcoord <- function(ddmm.) {
  dd <- floor(ddmm./100)
  mm. <- ddmm. - 100*dd
  dd. <- dd + mm./60
  return(dd.)
}
op$long <- -fixcoord(op$long)
op$lat <- fixcoord(op$lat)
op$opdate <- as.Date(op$opdate, "%d-%b-%y")
sel <- !apply(is.na(op[, c("lat", "long")]), 1, any) & op$year %in% 2006:2010
op <- op[sel, c("opid", "year", "vessel", "cruise", "serial", "port", "opdate", "time", "lat", "long", "surftemp")]

trop <- read.csv(paste0(datadir, "HURON_TROP.csv"), as.is=TRUE, 
  col.names=c("opid", "towtime", "fishtemp", "depth", "trdesign"))
trop$fishtemp <- as.numeric(trop$fishtemp)

optrop <- merge(op, trop)
optrop$portf <- as.factor(optrop$port)

cat <- read.csv(paste0(datadir, "HURON_TRCATCH.csv"), as.is=TRUE, 
  col.names=c("opid", "stage", "species", "n", "weight", "lfn"))
numvars <- c("stage", "weight", "lfn")
cat[, numvars] <- apply(cat[, numvars], 2, as.numeric)
bigcat <- merge(optrop, cat)
```



# DataTables - Tabular data display  

**DataTables** displays R matrices or data frames as interactive HTML tables that support filtering, pagination, and sorting.

```{r}
library(DT)
selvars <- c("opid", "year", "vessel", "cruise", "serial", "port", "opdate", 
  "fishtemp", "depth")
datatable(optrop[, selvars], options=list(pageLength=5), rownames=FALSE, 
  caption="Example DataTables", filter="top")
```



# MetricsGraphics - Scatterplots and line charts with D3  

MetricsGraphics enables easy creation of D3 scatterplots, line charts, and histograms.

```{r}
library(metricsgraphics)
library(htmlwidgets)
library(htmltools)
library(dplyr)

mjs_plot(optrop, x=depth, y=fishtemp) %>%
  mjs_point(color_accessor=portf, size_accessor=surftemp, 
    size_range=c(2, 15), color_type="category") %>%
  mjs_labs(x="Bottom depth  (m)", y="Fishing temperature  (C)")

wh <- 500
mjs_plot(optrop, x=depth, y=fishtemp, linked=TRUE, width=wh, height=wh, left=20, bottom=20) %>%
  mjs_point(color_accessor=surftemp, size_accessor=surftemp, 
    size_range=c(2, 15), color_type="category") %>%
  mjs_labs(x="Bottom depth  (m)", y="Fishing temperature  (C)") -> plot1

mjs_plot(optrop, x=portf, y=year, linked=TRUE, width=wh, height=wh) %>%
  mjs_axis_y(min_y=2006, max_y=2010) %>%
  mjs_point(color_accessor=surftemp, color_range=c("lightgray", "darkgray"), size_accessor=surftemp, 
    size_range=c(2, 15), color_type="category") %>%
  mjs_labs(x="Port code", y="Year") -> plot2

mjs_grid(plot1, plot2, ncol=2, nrow=1)
```



# dygraphs - Time series charting  

Dygraphs provides rich facilities for charting time-series data in R and includes support for many interactive features including series/point highlighting, zooming, and panning.



# Leaflet - Geo-spatial mapping  

Leaflet is a JavaScript library for creating dynamic maps that support panning and zooming along with various annotations like markers, polygons, and popups.

```{r}
library(leaflet)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

Make the map bigger.

```{r}
leaflet(optrop, width=1000, height=600) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat)
```

Color the symbols by depth.

```{r}
numpal <- colorNumeric("Blues", NULL)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth))
```

Add a legend.

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~numpal(depth)) %>%
  addLegend("bottomleft", pal=numpal, values=~depth, opacity=1,
    title="Bottom depth  (m)")
```

Highlight one year.

```{r}
facpal <- colorFactor(c("blue", "orange"), NULL)
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, color=~facpal(year==2010))
```

Provide a toggle for that one year.

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long[year!=2010], lat=~lat[year!=2010], 
    group="2006-2009") %>% 
  addCircleMarkers(lng=~long[year==2010], lat=~lat[year==2010], 
    group="2010") %>%   
  addLayersControl(overlayGroups=c("2006-2009", "2010"),
    options=layersControlOptions(collapsed=FALSE))
```

Show the opid when you click on a point.

```{r}
leaflet(optrop) %>% 
  addTiles() %>%
  addCircleMarkers(lng=~long, lat=~lat, popup=~htmlEscape(as.character(opid)))
```

Use a different background map.

```{r}
leaflet(optrop) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(lng=~long, lat=~lat)
```




# networkD3 - Graph data visualization with D3  

**networkD3** provides tools for creating D3 JavaScript network graphs from R.

```{r}
tot <- with(bigcat, tapply(n, list(opid, species), sum))

co.occur.d <- dist(!is.na(t(tot)), method="binary")
co.occur <- as.matrix(1 - co.occur.d)
co.occur[upper.tri(co.occur, diag=TRUE)] <- NA

library(reshape2)
mylinks <- melt(co.occur, na.rm=TRUE)
uniq.sp <- colnames(tot)
mylinks$source <- match(mylinks$Var1, uniq.sp) - 1
mylinks$target <- match(mylinks$Var2, uniq.sp) - 1
library(plotrix)
mylinks$width <- rescale(mylinks$value, c(0.1, 20))
mylinks <- mylinks[mylinks$value > 0.2, ]

mynodes <- data.frame(name=as.numeric(uniq.sp))
mynodes$group <- floor(mynodes$name/100)
mynodes$freq <- rescale(apply(tot, 2, sum, na.rm=TRUE), c(0.1, 30))

library(networkD3)
data(MisLinks, MisNodes)
forceNetwork(Links=mylinks, Nodes=mynodes, Source="source", Target="target", 
  Value="width", NodeID="name", Nodesize="freq", Group="group", opacity=0.9, 
  fontSize=30, linkDistance=JS('function(d) {', 'return 10*d.value;', '}'))
```

We can also plot dendrograms.

```{r}
clusdat <- t(tot)
clusdat[is.na(clusdat)] <- 0
hc <- hclust(dist(clusdat))
dendroNetwork(hc, fontSize=15, zoom=TRUE)
```


# d3heatmap - Interactive heatmaps with D3  

Interactive heatmaps with D3 including support for row/column highlighting and zooming.


```{r}
pd <- with(bigcat, tapply(n, list(port + depth/1000, species), mean))
pd[is.na(pd)] <- 0

library(d3heatmap)
d3heatmap(pd, scale="column", colors="Blues",
  xaxis_font_size="16pt", yaxis_font_size="16pt")
```


