---
title: "Recent Advances in Interactive Visualizations in R"
author: "Jean V. Adams"
date: "11 Jan 2016"
output:
  html_document: default
subtitle: USGS-GLSC All Hands Meeting
---

```{r setup, echo=FALSE} 
library(knitr) 
opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

# define two variables to keep track of Table and Figure numbers
tabcount <- 1
figcount <- 1
```

# **htmlwidgets** for R

> HTML widgets work just like R plots except they produce interactive web visualizations. 

> A line or two of R code is all it takes to produce a D3 graphic or Leaflet map. 

> HTML widgets can be used at the R console as well as embedded in R Markdown reports and Shiny web applications.

**DataTables** - Tabular data display   
**MetricsGraphics** - Scatterplots and line charts with D3  
**dygraphs** - Time series charting  
**Leaflet** - Geo-spatial mapping  
**networkD3** - Graph data visualization with D3  
**d3heatmap** - Interactive heatmaps with D3  

I will use historical data from the Lake Huron bottom trawl survey to demonstrate the capabilities of **htmlwidgets** for R.

For more information, visit the **htmlwidgets** showcase at http://www.htmlwidgets.org/showcase_networkD3.html.

# Read in the data

```{r, echo=FALSE}
setwd("C:/JVA/GitHub/vizzy")
datadir <- "C:/JVA/Lakes Vessels/Huron/2012/"
```

```{r}
op <- read.csv(paste0(datadir, "HURON_OP.csv"), as.is=TRUE, 
  col.names=c("opid", "year", "vessel", "serial", "lake", "port", "cruise", 
    "opdate", "time", "lat", "long", "surftemp", "weather", "windspeed", 
    "seacond", "winddirect", "distance"))
numvars <- c("lat", "long", "surftemp", "weather", "windspeed", "seacond", 
  "winddirect", "distance")
op[, numvars] <- apply(op[, numvars], 2, as.numeric)
fixcoord <- function(ddmm.) {
  dd <- floor(ddmm./100)
  mm. <- ddmm. - 100*dd
  dd. <- dd + mm./60
  return(dd.)
}
op$long <- -fixcoord(op$long)
op$lat <- fixcoord(op$lat)
op$opdate <- as.Date(op$opdate, "%d-%b-%y")
sel <- !apply(is.na(op[, c("lat", "long")]), 1, any)
op <- op[sel, c("opid", "year", "vessel", "cruise", "serial", "port", "opdate", "time", "lat", "long", "surftemp")]

trop <- read.csv(paste0(datadir, "HURON_TROP.csv"), as.is=TRUE, 
  col.names=c("opid", "towtime", "fishtemp", "depth", "trdesign"))
trop$fishtemp <- as.numeric(trop$fishtemp)

optrop <- merge(op, trop)
optrop$portf <- as.factor(optrop$port)

cat <- read.csv(paste0(datadir, "HURON_TRCATCH.csv"), as.is=TRUE, 
  col.names=c("opid", "stage", "species", "n", "weight", "lfn"))
numvars <- c("stage", "weight", "lfn")
cat[, numvars] <- apply(cat[, numvars], 2, as.numeric)
```

# DataTables - Tabular data display  

**DataTables** displays R matrices or data frames as interactive HTML tables that support filtering, pagination, and sorting.

```{r}
library(DT)
selvars <- c("opid", "year", "vessel", "cruise", "serial", "port", "opdate", 
  "fishtemp", "depth")
datatable(optrop[, selvars], options=list(pageLength=5), rownames=FALSE, 
  caption="Example DataTables", filter="top")
```

# MetricsGraphics - Scatterplots and line charts with D3  

MetricsGraphics enables easy creation of D3 scatterplots, line charts, and histograms.

```{r}
library(metricsgraphics)
library(htmlwidgets)
library(htmltools)
library(dplyr)

mjs_plot(optrop, x=depth, y=fishtemp) %>%
  mjs_point(color_accessor=portf, size_accessor=surftemp, 
    color_type="category") %>%
  mjs_labs(x="Bottom depth  (m)", y="Fishing temperature  (C)")

wh <- 500
mjs_plot(optrop, x=depth, y=fishtemp, linked=TRUE, width=wh, height=wh, left=20, bottom=20) %>%
  mjs_point(color_accessor=surftemp, size_accessor=surftemp, 
    size_range=c(1, 10), color_type="category") %>%
  mjs_labs(x="Bottom depth  (m)", y="Fishing temperature  (C)") -> plot1

mjs_plot(optrop, x=portf, y=year, linked=TRUE, width=wh, height=wh) %>%
  mjs_axis_y(min_y=1976, max_y=2012) %>%
  mjs_point(color_accessor=surftemp, size_accessor=surftemp, 
    size_range=c(1, 10), color_type="category") %>%
  mjs_labs(x="Port code", y="Year") -> plot2

mjs_grid(plot1, plot2, ncol=2, nrow=1)
```

# dygraphs - Time series charting  

Dygraphs provides rich facilities for charting time-series data in R and includes support for many interactive features including series/point highlighting, zooming, and panning.



# Leaflet - Geo-spatial mapping  

Leaflet is a JavaScript library for creating dynamic maps that support panning and zooming along with various annotations like markers, polygons, and popups.

```{r}
# convert your data frame to a spatial data frame
xy <- optrop[, c("long", "lat")]
rest <- optrop[, !(names(optrop) %in% c("long", "lat"))]

library(sp)
spdf <- SpatialPointsDataFrame(coords=xy, data=rest,
  proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

library(leaflet)
pal <- colorNumeric("YlOrRd", NULL)
leaflet(spdf, width=910, height=500) %>% 
  addTiles() %>%
  addCircleMarkers(color = ~pal(port))
```


# networkD3 - Graph data visualization with D3  



# d3heatmap - Interactive heatmaps with D3  





# Tables

You can create tables using simple source code (Table `r tabcount`) or using the *knitr* function *kable* (Table `r tabcount + 1`).

**Table `r tabcount`.** First few rows of *mtcars* using simple source code.
```{r, comment=NA, echo=FALSE}
head(mtcars)
tabcount <- tabcount + 1
```

**Table `r tabcount`.** First few rows of *mtcars* using *kable*.
```{r, results='asis', echo=FALSE}
knitr::kable(head(mtcars))
tabcount <- tabcount + 1
```

# Figures

You create figures, either using the default size (Figure `r figcount`) or a custom size (Figure `r figcount + 1`).



```{r}
par(mar=c(4, 4, 1, 1), las=1)
plot(cars)
```

**Figure `r figcount`.** Simple plot of *cars*, default size.

```{r, echo=FALSE}
figcount <- figcount + 1
```



```{r, fig.width=3, fig.height=3}
par(mar=c(4, 4, 1, 1), las=1)
plot(cars)
```

**Figure `r figcount`.** Simple plot of *cars*, custom size.

```{r, echo=FALSE}
figcount <- figcount + 1
```



# Miscellany

## Header fonts 
### Header 3 
#### Header 4 
##### Header 5 
###### Header 6 

# Lists

Unordered List:

* This item
* That item
    + A that subitem

Ordered List:

1. Item 1
    a. Sub item a
    a. Sub item b
1. Item 2
    
You can write [equations](https://support.rstudio.com/hc/en-us/articles/200486328-Equations-in-R-Markdown), either inline, $A = \pi*r^{2}$, or centered,
$$\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_{i}.$$

You can write **bold**, *italics*, superscripts, R^2^, subscripts, X~ij~.  You can ~~cross things out~~.  And you can draw a horizontal line. 

*** 

> Block quote for a long piece of indented text. When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
